---
title: "R-Tutorials Using Titanic Data"
author: "Josemari Feliciano"
date: "7/10/2017"
output:
  pdf_document: default
---

# There are three major parts to this tutorials:

1.  How to read and extract data from RData file
2.  How to create basic graphs 
3.  How to run basic descriptive statistics
4.  How to run and read results of inferential statistics

### Tutorial Notes:  

The Beauty of R Studio is the ability to have notebooks for data analysis.  It allows us to run R-code inline similar to IPython Notebooks.  

Think of it as a fancy chemistry notebook where you can somehow run the experiments itself in the notebook -- but for data analysis!  For this notebook and R tutorial, I will use the titanic data for analysis.  This assumes that you have R and RStudio loaded.  


__Note__ :  mosaic package is required for this if you want to run data yourself and not to simply view them.   If you do not have this, run this code in your r-console independently:  

```
install.packages("ggplot2")
``` 



  
  
## Part 1:  Reading data  

We probably want to load the titanic data first. We probably want to load the mosaic library out of the way.  

### Loading Data:
We can accomplish both data loading and library call with the following script:

```{r, message=FALSE, warning=FALSE}
load("Titanic.Rdata")
library("mosaic")
```
  
### Which Variables Are In Data?  
Now that the data has been loaded, we probably want to see which data variables we will deal with!  Below, we will use `names()` to print the variable names in our data to have them handy.  

```{r}
names(Titanic)
```

The `names()` function we just ran displayed the 6 variables within the Titanic Data which include Gender, Age, Name, Fare, Class and Survived.   
  
Of course, you could have looked at the actual CSV file or RData file directly. But functions like `names()` are very useful when wrangling data such as JSON and similar data types.  
  
### Exploring Data Types:

The `sapply()` function is very useful for this.  The data types do make sense. 
  
```{r}
sapply(Titanic,class)
```

The output above does make sense; data types seem to match what we would expect.  Age is numeric.  And Gender is a 'factor', commonly known as string in other computer langauges. 

Another function we could have used is `str()`, but the output can be messy and dense. `str()` does provide more information on the variables in our data.  

```{r}
str(Titanic)
```

As we can see, more information are provided. For instance, the output shows that there are 2 levels for Gender, "Famale" and "Male".  But `str()` can look a bit messy.  

_Note:_ We could have used `View(Titanic)` to open the spreadsheet in a difference pane or tab in RStudio.  But that will require sifting through the spreadsheet.  It's good to familiarize with both `str()` and `sapply()` functions so we don't have leave our tab.  

## Creating Basic Graphs

There are so many ways to create graphs and the resources and synthax for R can be overwhelming.  So I will demonstrate a few ways to depict our data in a meaningful way.

### Comparing Genders:



