---
title: "Tutorials Of R Using Titanic Data"
author: "Josemari Feliciano"
date: "7/10/2017"
output:
  pdf_document: default
---


# Tutorial Notes:  

The Beauty of R Studio is the ability to have notebooks for data analysis.  It allows us to run r-code inline.  Think of it as a fancy chemistry notebook where you can run the the experiments in the notebook itself -- but for data analysis!  For this notebook, I will use the titanic data for analysis.  This assumes that you have R and RStudio loaded.  



__Note__ :  mosaic package is required for this if you want to run data yourself and not to simply view them.   If you do not have this, run this code in your r-console independently:

```
install.packages("ggplot2")
```


Getting started:  We probably want to load the titanic data first. We probably want to load the mosaic library out of the way.  So we can accomplish both with the following script:

```{r, message=FALSE, warning=FALSE}
load("Titanic.Rdata")
library("mosaic")
```

Now that the data has been loaded, we probably want to see which data variables we will deal with!  











```{r}
summary(Titanic)
```

```{r}
bargraph( ~ Class, data =Titanic)
```

